Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-30 paper/main.tex
updates to conclusion
removed stray comments, added an mbox

IPython preprocesses all executed code as plain text before it is sent
to the Python compiler, in order to handle special constructs like
\mbox{\texttt{!}-prefixed} shell commands and \mbox{\texttt{\%}-pre}\-fixed ``magic'' commands. Davos uses
this same process to invisibly transform \texttt{smuggle} statements into
syntactically valid Python code. The Davos parser uses a
regular expression to match lines of code containing \texttt{smuggle}
statements (and, optionally, onion comments), extract relevant
information from their text, and replace them with equivalent calls to
the \texttt{smuggle()} function. For example, if a user runs a
notebook cell containing
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet6}
\end{center}
the code that is actually executed by the Python interpreter would be
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet7}
\end{center}
The call to the \texttt{smuggle()} function carries out
Davos's central logic by determining whether the smuggled
package must be installed, carrying out the installation if necessary,
and subsequently loading it into the namespace. This process is
outlined in Figure~\ref{fig:flow-chart}. Because the
\texttt{smuggle()} function is defined in the notebook namespace, it
is also possible (though never necessary) to call it
directly. Deactivating Davos will delete the name
``\texttt{smuggle}'' from the namespace, unless its value has been
overwritten and no longer refers to the \texttt{smuggle()}
function. It will also deregister the Davos parser from the
set of input transformers run when each notebook cell is
executed.

@@

%The example code throughout Section \ref{subsec:onion} illustrates how Davos is most typically used:
%By including a series of \texttt{smuggle} statements and onion comments with version specifiers or other options in an IPython notebook, researchers can share their code and its dependencies in a single file that can be easily run without any additional tools or setup, creates and manages its own isolated environment, automatically installs its required packages at runtime, and ensures that the package versions with which it is run do not change unexpectedly.
The example code throughout Section \ref{subsec:onion} illustrates a typical use case that we envision for Davos: a series of smuggle statements and onion comments with version specifiers or other options collectively describes and automatically constructs a reproducible environment for running the code that follows it.
When added to the top of a Jupyter notebook, this allows researchers to bundle their code and its dependencies into a single file that can be easily shared and run without any additional tools or setup, automatically installs its required packages at runtime, isolates them from the user's main Python environment, and ensures their versions do not change unexpectedly over time.
In this section, we have contrived a more complex scenario to highlight some of Davos's more advanced features, and illustrate how they may be used to handle certain challenges that can arise when writing, running, and sharing reproducible scientific code.

@@

The example provided in Figure~\ref{fig:illustrative-example} demonstrates how
Davos can be used to circumvent these incompatibilities by
temporarily switching between different versions of the same package within a single runtime.
%carefully controlling which versions of each package are used in different parts of the notebook.
The example shows how a dataset and model that require
now-incompatible components of the \texttt{pandas} and
\texttt{scikit-learn}~\cite{PedrEtal11} libraries can be loaded in (using older
versions of each package) and used alongside more recent versions of each
package that provide new and improved functionality. When included at the top
of a Jupyter notebook, the code in Figure~\ref{fig:illustrative-example}
ensures that these objects will be loaded successfully and analyzed using the
same set of package versions no matter when or by whom the notebook is run.

After installing and importing Davos (lines 1--2), we first use the \texttt{davos.require\_\-python()} function to constrain the Python version used to run the notebook (see Sec.~\ref{subsec:toplevel}).
As described above, the example code in Figure \ref{fig:illustrative-example} loads two different versions of the \texttt{pandas} library: first, an older version needed to access a dataset saved in an outmoded format, then a newer one to use throughout the remainder of the notebook.
We therefore want to make sure upfront (in line 6) that the notebook's Python version falls within the range of versions that both of these two versions of \texttt{pandas} support.
%Line 6 therefore ensures upfront that the notebook's Python version falls within the overlap between the ranges of Python versions that these two versions of \texttt{pandas} support.
If it does not, the function in line 6 will raise an error that includes a message to this effect (lines 4--5).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}

@@

Line 11 then enables
Davos's \texttt{auto\_rerun} option (see Sec.~\ref{subsec:config}) before smuggling the next
two packages: \texttt{NumPy} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions specified in their onion comments may
require restarting the notebook kernel if different versions were previously
imported during the same interpreter session---including internally by other packages.
Enabling \texttt{auto\_rerun} allows Davos to handle kernel restarts automatically and continue running the code seamlessly without user intervention.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}
In the case of \texttt{NumPy}, whether or not a kernel restart is necessary will depend on the user's existing Python environment.
The \texttt{joblib} package has an optional dependency on \texttt{NumPy} for memoizing and parallelizing array operations, and will \texttt{import numpy} internally to enable these features if the package is available.
If the user already has \texttt{NumPy} installed in their Python environment when \texttt{joblib} is smuggled in line 9, their installed version is different from the one specified in the onion comment on line 12, and there were changes made to \texttt{NumPy}'s C extensions between those two versions, then Davos will automatically restart the kernel and re-run the lines above.
The newly smuggled version would then be used both in the notebook itself and by \texttt{joblib} internally.
% (Note that outside the context of an illustrative example, one could avoid a kernel restart here altogether simply by smuggling \texttt{NumPy} before \texttt{joblib}.)

@@

In lines 30--37, we demonstrate another aspect of Davos's functionality that supports more advanced installation scenarios.
%The \texttt{ipywidgets}~\cite{FredEtal15} package (also known as Jupyter Widgets) provides a Python API for creating various JavaScript widgets within a notebook, and the \texttt{widgetsnbextension} package provides the JavaScript machinery needed by the the notebook frontend to display them.
The \texttt{ipywidgets}~\cite{FredEtal15} package (also known as Jupyter Widgets) provides a Python API for creating interactive JavaScript widgets within a notebook.
It depends on the \texttt{widgetsnbextension} package, which provides the JavaScript machinery needed by the notebook frontend to display these widgets.
%A complication is that, while \texttt{ipywidgets} must be installed in a location that is accessible from the IPython kernel (i.e., the Python runtime of the notebook itself), \texttt{widgetsnbextension} must be made accessible to the Jupyter notebook server, which is a separate Python runtime.
%A complication is that, while \texttt{ipywidgets} must be installed in an environment accessible from the IPython kernel (i.e., importable into the notebook itself), \texttt{widgetsnbextension} must be installed in the environment that houses the Jupyter notebook server.
A complication is that \texttt{ipywidgets} must be installed in a location that is accessible from the IPython kernel (i.e., the Python runtime within the notebook itself), while \texttt{widgetsnbextension} must be installed in the environment that houses the Jupyter notebook server (a separate Python runtime that serves and manages the notebook frontend client).
In many basic setups, the IPython kernel and notebook server exist in the same environment.
However, a common ``advanced'' approach entails running the notebook server from a base environment, with additional environments each providing their own separate, interchangeable IPython kernels.

Lines 30--37 account for both of these possibilities programmatically:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
First, in line 30, we set the \texttt{davos.project} attribute to \texttt{None} to temporarily
%disable Davos's project isolation system and
allow installing smuggled packages outside of the notebook's project directory.
As noted in Section \ref{subsec:projects}, this is typically discouraged, as doing so can risk interfering with the user's Python environment if existing package versions are overwritten.
In this particular case, however, a combination of factors make this relatively safe and inconsequential.
First, the package we need to install directly into the notebook server environment (\texttt{widgetsnbextension}) is smuggled without an accompanying onion comment (line 34), meaning that Davos will not replace any version the user may already have installed.
Second, the package has no dependencies of its own, so if Davos does install it, no other packages will be installed or updated as a side effect.
Third, the package itself provides no functionality outside of rendering Jupyter widgets, so its presence would not alter any other code's expected behavior.

@@

\todo{mention notebook reproducibility, cell order, multiple package versions re: reviewer's comment; note importance of running lines 14--19 \& 38--40 in single cell}

@@

The Davos package supports reproducible research by providing
a novel, lightweight system for sharing notebook-based code. But
perhaps the most exciting uses of the Davos package are those
that we have \textit{not} yet considered or imagined. We hope that the
research and scientific Python communities will find Davos to provide a convenient
means of managing project dependencies to facilitate code sharing and collaboration. We
also hope that some of the more advanced applications of our package
might lead to new insights or discoveries.

>>>>

IPython preprocesses all executed code as plain text before it is sent
to the Python compiler, in order to handle special constructs like
\mbox{\texttt{!}-pre}\-fixed shell commands and \mbox{\texttt{\%}-pre}\-fixed ``magic'' commands. Davos uses
this same process to invisibly transform \texttt{smuggle} statements into
syntactically valid Python code. The Davos parser uses a
regular expression to match lines of code containing \texttt{smuggle}
statements (and, optionally, onion comments), extract relevant
information from their text, and replace them with equivalent calls to
the \texttt{smuggle()} function. For example, if a user runs a
notebook cell containing
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet6}
\end{center}
the code that is actually executed by the Python interpreter would be
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet7}
\end{center}
The call to the \texttt{smuggle()} function carries out
Davos's central logic by determining whether the smuggled
package must be installed, carrying out the installation if necessary,
and subsequently loading it into the namespace. This process is
outlined in Figure~\ref{fig:flow-chart}. Because the
\texttt{smuggle()} function is defined in the notebook namespace, it
is also possible (though never necessary) to call it
directly. Deactivating Davos will delete the name
``\texttt{smuggle}'' from the namespace, unless its value has been
overwritten and no longer refers to the \texttt{smuggle()}
function. It will also deregister the Davos parser from the
set of input transformers run when each notebook cell is
executed.

@@

The example code throughout Section \ref{subsec:onion} illustrates a typical use case that we envision for Davos: a series of smuggle statements and onion comments with version specifiers or other options collectively describes and automatically constructs a reproducible environment for running the code that follows it.
When added to the top of a Jupyter notebook, this allows researchers to bundle their code and its dependencies into a single file that can be easily shared and run without any additional tools or setup, automatically installs its required packages at runtime, isolates them from the user's main Python environment, and ensures their versions do not change unexpectedly over time.
In this section, we have contrived a more complex scenario to highlight some of Davos's more advanced features, and illustrate how they may be used to handle certain challenges that can arise when writing, running, and sharing reproducible scientific code.

@@

The example provided in Figure~\ref{fig:illustrative-example} demonstrates how
Davos can be used to circumvent these incompatibilities by
temporarily switching between different versions of the same package within a single runtime.
The example shows how a dataset and model that require
now-incompatible components of the \texttt{pandas} and
\texttt{scikit-learn}~\cite{PedrEtal11} libraries can be loaded in (using older
versions of each package) and used alongside more recent versions of each
package that provide new and improved functionality. When included at the top
of a Jupyter notebook, the code in Figure~\ref{fig:illustrative-example}
ensures that these objects will be loaded successfully and analyzed using the
same set of package versions no matter when or by whom the notebook is run.

After installing and importing Davos (lines 1--2), we first use the \texttt{davos.require\_\-python()} function to constrain the Python version used to run the notebook (see Sec.~\ref{subsec:toplevel}).
As described above, the example code in Figure \ref{fig:illustrative-example} loads two different versions of the \texttt{pandas} library: first, an older version needed to access a dataset saved in an outmoded format, then a newer one to use throughout the remainder of the notebook.
We therefore want to make sure upfront (in line 6) that the notebook's Python version falls within the range of versions that both of these two versions of \texttt{pandas} support.
If it does not, the function in line 6 will raise an error that includes a message to this effect (lines 4--5).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}

@@

Line 11 then enables
Davos's \texttt{auto\_rerun} option (see Sec.~\ref{subsec:config}) before smuggling the next
two packages: \texttt{NumPy} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions specified in their onion comments may
require restarting the notebook kernel if different versions were previously
imported during the same interpreter session---including internally by other packages.
Enabling \texttt{auto\_rerun} allows Davos to handle kernel restarts automatically and continue running the code seamlessly without user intervention.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}
In the case of \texttt{NumPy}, whether or not a kernel restart is necessary will depend on the user's existing Python environment.
The \texttt{joblib} package has an optional dependency on \texttt{NumPy} for memoizing and parallelizing array operations, and will \texttt{import numpy} internally to enable these features if the package is available.
If the user already has \texttt{NumPy} installed in their Python environment when \texttt{joblib} is smuggled in line 9, their installed version is different from the one specified in the onion comment on line 12, and there were changes made to \texttt{NumPy}'s C extensions between those two versions, then Davos will automatically restart the kernel and re-run the lines above.
The newly smuggled version would then be used both in the notebook itself and by \texttt{joblib} internally.

@@

In lines 30--37, we demonstrate another aspect of Davos's functionality that supports more advanced installation scenarios.
The \texttt{ipywidgets}~\cite{FredEtal15} package (also known as Jupyter Widgets) provides a Python API for creating interactive JavaScript widgets within a notebook.
It depends on the \texttt{widgetsnbextension} package, which provides the JavaScript machinery needed by the notebook frontend to display these widgets.
A complication is that \texttt{ipywidgets} must be installed in a location that is accessible from the IPython kernel (i.e., the Python runtime within the notebook itself), while \texttt{widgetsnbextension} must be installed in the environment that houses the Jupyter notebook server (a separate Python runtime that serves and manages the notebook frontend client).
In many basic setups, the IPython kernel and notebook server exist in the same environment.
However, a common ``advanced'' approach entails running the notebook server from a base environment, with additional environments each providing their own separate, interchangeable IPython kernels.

Lines 30--37 account for both of these possibilities programmatically:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
First, in line 30, we set the \texttt{davos.project} attribute to \texttt{None} to temporarily
allow installing smuggled packages outside of the notebook's project directory.
As noted in Section \ref{subsec:projects}, this is typically discouraged, as doing so can risk interfering with the user's Python environment if existing package versions are overwritten.
In this particular case, however, a combination of factors make this relatively safe and inconsequential.
First, the package we need to install directly into the notebook server environment (\texttt{widgetsnbextension}) is smuggled without an accompanying onion comment (line 34), meaning that Davos will not replace any version the user may already have installed.
Second, the package has no dependencies of its own, so if Davos does install it, no other packages will be installed or updated as a side effect.
Third, the package itself provides no functionality outside of rendering Jupyter widgets, so its presence would not alter any other code's expected behavior.

@@

@@

The Davos package supports reproducible research by providing
a novel, lightweight system for sharing notebook-based code. It
expands on Python's ``batteries included'' philosophy to enable running shared
notebooks with no setup required, and defines a simple, self-documenting format for
specifying dependencies in the spirit of ``literate programming'' paradigm Jupyter notebooks promote.
But perhaps the most exciting uses of the Davos package are those
that we have \textit{not} yet considered or imagined. We hope that the
research and scientific Python communities will find Davos to provide a convenient
means of managing project dependencies to facilitate code sharing and collaboration. We
also hope that some of the more advanced applications of our package
might lead to new insights or discoveries.
